---
title: "AMOD5430 Project"
author: "Hamza Chaudhry"
date: "2025-07-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(ggplot2)

setwd("~/Desktop/AMOD5430")

# Netflix dataset im using
netflix <- read.csv("netflix_titles.csv", stringsAsFactors = FALSE)

# View column names and sample data
head(netflix)
str(netflix)
```

# Data Cleaning and Preparation

```{r}
# Using the date format
netflix$date_added <- as.Date(netflix$date_added, format = "%B %d, %Y")

netflix$year_added <- format(netflix$date_added, "%Y")

# Cleaning the country and listed_in columns 
netflix$country[is.na(netflix$country)] <- "Unknown"
netflix$listed_in <- as.factor(netflix$listed_in)
netflix$type <- as.factor(netflix$type)

```

# Figure 1: The Number of titles added to Netflix overtime
```{r}
ggplot(netflix, aes(x = year_added)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Figure 1: Number of Titles Added by Year",
       x = "Year Added", y = "Count of Titles") +
  theme_minimal()
```
# Figure 1 shows the number of titles added to Netflix each year based on the date_added column. 
# There’s a clear surge from 2016 to 2020, suggesting Netflix aggressively expanded its library during this period. 
# The decline post-2020 may reflect strategic shifts or disruptions during the pandemic. 
# A small number of titles have missing dates, shown as “NA.”

# Figure 2: Movies vs TV shows on Netflix
```{r}
ggplot(netflix, aes(x = type, fill = type)) +
  geom_bar() +
  labs(title = "Figure 2: Distribution of Content Type",
       x = "Type", y = "Count") +
  theme_minimal()
```
# Figure 2 illustrates the overall distribution of content types on Netflix. 
# Movies dominate the platform, with more than twice the count of TV Shows. 
# This shows Netflix’s preference for feature-length content, but serialized content is also important for user retention

# Figure 3:Top 10 countries producing content on Netflix
```{r}

top_countries <- sort(table(netflix$country), decreasing = TRUE)[1:10]
barplot(top_countries,
        col = "darkred",
        main = "Figure 3: Top 10 Countries by Number of Titles",
        ylab = "Count of Titles",
        las = 2)
```
# Figure 3 shows the top 10 countries producing content on Netflix. 
# The United States leads by a wide margin, followed by India and the UK. 
# This indicates that while Netflix has global reach, much of its content still originates from English-speaking or major media markets.

# Figure 4: Top genres on Netflix
```{r}

genre_counts <- table(netflix$listed_in)
top_genres <- sort(genre_counts, decreasing = TRUE)[1:10]

top_genres_df <- as.data.frame(top_genres)
colnames(top_genres_df) <- c("Genre", "Count")

library(ggplot2)
ggplot(top_genres_df, aes(x = reorder(Genre, Count), y = Count)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Figure 4: Top 10 Genres",
       x = "Genre", y = "Number of Titles") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))



```
# Figure 4 highlights the most common genres on Netflix. 
# International movies, documentaries, and stand-up comedy dominate the list. 
# The presence of multiple types of “International” genres shows Netflix's global reach 
# Family-friendly and kid-oriented content also reinforce the previous point.

# Figure 5: Movie durations on Netflix
```{r}

movie_data <- subset(netflix, type == "Movie")
movie_data$duration_mins <- as.numeric(gsub(" min", "", movie_data$duration))

hist(movie_data$duration_mins,
     main = "Figure 5: Movie Duration Distribution",
     xlab = "Duration (minutes)",
     col = "skyblue",
     breaks = 30)
```
# Figure 5 presents the distribution of movie durations. 
# Most films on Netflix fall in the 80–110 minute range, which is standard feature film length. 
# Very short and very long movies are rare, which is as expected.

# --- Conclusion ---
# Through our analysis, we found several trends in Netflix content.
# - The number of titles added grew steadily, peaking in recent years.
# - Movies dominate the platform compared to TV Shows.
# - The United States leads in content production, but Netflix hosts titles from many countries.
# - Drama and international genres are among the most common.
# - Movie durations cluster around 90-100 minutes, with a few outliers.

# These insights are important for content creators to understand Netflix's strategy and also to see what works and what does not.