---
title: "Modeling Geographic Disparities in COVID-19 Vaccine Uptake"
author: "Hamza Chaudhry"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    fig-width: 7
    fig-height: 5
    fig-dpi: 300
    embed-resources: true
    code-fold: true
    code-tools: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 6.5
    fig-height: 4.5
    fig-dpi: 300
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
execute:
  echo: true
  warning: false
  message: false
  cache: true
bibliography: references.bib
---

```{r}
#| label: setup
#| cache: false

# =============================================================================
# LIBRARIES
# =============================================================================

library(tidyverse)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(scales)
library(ggrepel)
library(corrplot)
library(factoextra)
library(cluster)
library(lmtest)
library(sandwich)
library(broom)
library(knitr)
library(kableExtra)
library(countrycode)
library(RColorBrewer)
library(patchwork)
library(car)

# =============================================================================
# GLOBAL THEME — applied to every ggplot in this document
# Follows lecture guidelines: sans-serif, no chartjunk, no background fill,
# axes readable horizontally, ~4:3 ratio
# =============================================================================

theme_thesis <- theme_classic(base_size = 11, base_family = "sans") +
  theme(
    plot.title       = element_text(size = 12, face = "bold"),
    axis.title       = element_text(size = 10),
    axis.text        = element_text(size = 9),
    legend.title     = element_text(size = 10),
    legend.text      = element_text(size = 9),
    strip.text       = element_text(size = 9, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

theme_set(theme_thesis)

# Continent colour palette — distinguishable in colour and B&W (via shape)
continent_colours <- c(
  "Africa"        = "#E69F00",
  "Asia"          = "#56B4E9",
  "Europe"        = "#009E73",
  "North America" = "#F0E442",
  "Oceania"       = "#0072B2",
  "South America" = "#D55E00"
)

continent_shapes <- c(16, 17, 15, 18, 8, 3)
```

# Introduction

Modern history saw one of the biggest global health interventions with the
rollout of COVID-19 vaccines. Despite widespread availability, vaccination
uptake varied enormously across countries and regions. Understanding what
drives these disparities — economic capacity, demographic structure, and
government policy — is essential for evaluating past responses and informing
future public health interventions.

This analysis uses the Our World in Data (OWID) global COVID-19 dataset to
examine geographic variation in vaccine uptake across approximately 200
countries. The primary outcome of interest is the percentage of the population
that is fully vaccinated. Explanatory variables include GDP per capita, the
Human Development Index, median age, the proportion of the population aged
65 and older, and a government stringency index.

# Data

## Loading the Dataset

```{r}
#| label: load-data

data_path <- "~/Desktop/AMOD5610/owid-covid-data.csv"

raw_data <- read_csv(data_path, show_col_types = FALSE)
```

The dataset contains **`r nrow(raw_data)`** rows and **`r ncol(raw_data)`**
columns, covering countries worldwide from early 2020 onward.

## Variable Selection

The following variables are extracted from the full dataset in line with the
study design outlined in the research proposal.

```{r}
#| label: select-vars

selected_vars <- c(
  "iso_code",
  "continent",
  "location",
  "date",
  "people_fully_vaccinated_per_hundred",
  "people_vaccinated_per_hundred",
  "total_boosters_per_hundred",
  "gdp_per_capita",
  "median_age",
  "aged_65_older",
  "human_development_index",
  "stringency_index",
  "population"
)

df <- raw_data %>%
  select(all_of(selected_vars))
```

## Data Cleaning

### Removing Aggregate Rows

OWID includes rows for aggregates such as "World", "Asia", and income group
bands. These are identified by ISO codes beginning with `OWID_` and are
removed to retain individual countries only.

```{r}
#| label: remove-aggregates

df <- df %>%
  filter(!str_starts(iso_code, "OWID_")) %>%
  filter(!is.na(continent)) %>%
  mutate(date = as.Date(date))
```

After filtering, the dataset covers **`r n_distinct(df$location)`** countries.

### Creating a Country-Level Snapshot

Since the primary interest is in overall vaccination outcomes rather than
daily changes, a single cross-sectional snapshot is constructed for each
country. For vaccination variables the most recent non-missing value is used;
for socioeconomic variables (which are essentially time-invariant in this
dataset) the last non-missing value is also used. Government stringency is
averaged across the entire pandemic period to capture overall policy intensity.

```{r}
#| label: snapshot

country_snapshot <- df %>%
  arrange(location, date) %>%
  group_by(iso_code, location, continent) %>%
  summarise(
    fully_vax_pct   = last(people_fully_vaccinated_per_hundred[!is.na(people_fully_vaccinated_per_hundred)]),
    one_dose_pct    = last(people_vaccinated_per_hundred[!is.na(people_vaccinated_per_hundred)]),
    booster_pct     = last(total_boosters_per_hundred[!is.na(total_boosters_per_hundred)]),
    gdp_per_capita  = last(gdp_per_capita[!is.na(gdp_per_capita)]),
    median_age      = last(median_age[!is.na(median_age)]),
    aged_65_older   = last(aged_65_older[!is.na(aged_65_older)]),
    hdi             = last(human_development_index[!is.na(human_development_index)]),
    stringency_mean = mean(stringency_index, na.rm = TRUE),
    population      = last(population[!is.na(population)]),
    last_vax_date   = max(date[!is.na(people_fully_vaccinated_per_hundred)],
                          na.rm = TRUE),
    .groups = "drop"
  )

country_clean <- country_snapshot %>%
  filter(!is.na(fully_vax_pct)) %>%
  mutate(
    log_gdp      = log(gdp_per_capita),
    income_group = countrycode(iso_code, origin = "iso3c",
                               destination = "wb.income",
                               warn = FALSE)
  )
```

The cleaned snapshot retains **`r nrow(country_clean)`** countries with valid
fully-vaccinated data.

### Missing Value Summary

```{r}
#| label: missing-values

missing_summary <- country_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing (n)") %>%
  filter(`Missing (n)` > 0) %>%
  arrange(desc(`Missing (n)`))

missing_summary %>%
  kable(caption = "Missing values per variable in the country-level snapshot.",
        booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, position = "left")
```

# Exploratory Data Analysis

## Distribution of Full Vaccination Rates

```{r}
#| label: fig-hist
#| fig-cap: "Distribution of full vaccination rates across countries. Each bar represents the number of countries falling within a given percentage range."

ggplot(country_clean, aes(x = fully_vax_pct)) +
  geom_histogram(bins = 25, fill = "grey40", colour = "white",
                 linewidth = 0.3) +
  labs(
    x = "Population fully vaccinated (%)",
    y = "Number of countries"
  )
```

## Vaccination Rates by Continent

```{r}
#| label: fig-dotplot-continent
#| fig-cap: "Mean full vaccination rate by continent. A dot plot is used in preference to a bar chart, as point position on a common axis is more accurately decoded (Cleveland & McGill, 1985)."

continent_summary <- country_clean %>%
  group_by(continent) %>%
  summarise(
    mean_vax   = mean(fully_vax_pct, na.rm = TRUE),
    median_vax = median(fully_vax_pct, na.rm = TRUE),
    n          = n(),
    .groups    = "drop"
  ) %>%
  arrange(mean_vax)

ggplot(continent_summary,
       aes(x = mean_vax, y = reorder(continent, mean_vax))) +
  geom_segment(aes(xend = 0, yend = reorder(continent, mean_vax)),
               colour = "grey70", linewidth = 0.5, linetype = "dashed") +
  geom_point(size = 4, colour = "grey20", shape = 16) +
  geom_text(aes(label = paste0(round(mean_vax, 1), "%")),
            hjust = -0.3, size = 3, family = "sans") +
  scale_x_continuous(limits = c(0, 90),
                     labels = label_percent(scale = 1)) +
  labs(
    x = "Mean population fully vaccinated (%)",
    y = NULL
  )
```

## Distribution Within Continents

```{r}
#| label: fig-boxplot
#| fig-cap: "Distribution of full vaccination rates by continent. Individual country values are overlaid as jittered points to show the underlying data."

ggplot(country_clean,
       aes(x = reorder(continent, fully_vax_pct, FUN = median),
           y = fully_vax_pct)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5,
               fill = "grey90", colour = "grey30", linewidth = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.2, colour = "grey40") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Population fully vaccinated (%)"
  )
```

## Vaccination Roll-out Over Time

```{r}
#| label: fig-timeseries
#| fig-cap: "Mean full vaccination rate by continent over time. Both colour and line type are used to distinguish continents, ensuring the figure is readable in black and white."
#| fig-width: 8
#| fig-height: 5

ts_data <- df %>%
  filter(!str_starts(iso_code, "OWID_"),
         !is.na(continent),
         !is.na(people_fully_vaccinated_per_hundred)) %>%
  group_by(continent, date) %>%
  summarise(mean_fully_vax = mean(people_fully_vaccinated_per_hundred,
                                  na.rm = TRUE),
            .groups = "drop")

ggplot(ts_data,
       aes(x = date, y = mean_fully_vax,
           colour = continent, linetype = continent)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = continent_colours, name = "Continent") +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted", "longdash", "twodash", "dotdash"),
    name   = "Continent"
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Date",
    y = "Mean population fully vaccinated (%)"
  ) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1, size = 8),
    legend.position = "right"
  )
```

## Relationship Between Key Predictors and Vaccination Rate

```{r}
#| label: fig-scatter-gdp
#| fig-cap: "Relationship between log GDP per capita and full vaccination rate. Both colour and point shape distinguish continents. The regression line is fitted across all countries."

ggplot(country_clean %>% filter(!is.na(gdp_per_capita)),
       aes(x = log_gdp, y = fully_vax_pct,
           colour = continent, shape = continent)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              linewidth = 0.8) +
  scale_colour_manual(values = continent_colours, name = "Continent") +
  scale_shape_manual(values = continent_shapes, name = "Continent") +
  scale_x_continuous(
    breaks = log(c(1000, 2500, 5000, 10000, 25000, 50000, 100000)),
    labels = c("$1k", "$2.5k", "$5k", "$10k", "$25k", "$50k", "$100k")
  ) +
  labs(
    x = "GDP per capita (log scale, USD)",
    y = "Population fully vaccinated (%)"
  ) +
  theme(legend.position = "right")
```

```{r}
#| label: fig-scatter-hdi
#| fig-cap: "Relationship between Human Development Index and full vaccination rate."

ggplot(country_clean %>% filter(!is.na(hdi)),
       aes(x = hdi, y = fully_vax_pct,
           colour = continent, shape = continent)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              linewidth = 0.8) +
  scale_colour_manual(values = continent_colours, name = "Continent") +
  scale_shape_manual(values = continent_shapes, name = "Continent") +
  labs(
    x = "Human Development Index",
    y = "Population fully vaccinated (%)"
  ) +
  theme(legend.position = "right")
```

```{r}
#| label: fig-scatter-age
#| fig-cap: "Relationship between median age and full vaccination rate."

ggplot(country_clean %>% filter(!is.na(median_age)),
       aes(x = median_age, y = fully_vax_pct,
           colour = continent, shape = continent)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              linewidth = 0.8) +
  scale_colour_manual(values = continent_colours, name = "Continent") +
  scale_shape_manual(values = continent_shapes, name = "Continent") +
  labs(
    x = "Median age (years)",
    y = "Population fully vaccinated (%)"
  ) +
  theme(legend.position = "right")
```

```{r}
#| label: fig-scatter-stringency
#| fig-cap: "Relationship between mean government stringency index and full vaccination rate."

ggplot(country_clean %>% filter(!is.na(stringency_mean)),
       aes(x = stringency_mean, y = fully_vax_pct,
           colour = continent, shape = continent)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              linewidth = 0.8) +
  scale_colour_manual(values = continent_colours, name = "Continent") +
  scale_shape_manual(values = continent_shapes, name = "Continent") +
  labs(
    x = "Mean government stringency index",
    y = "Population fully vaccinated (%)"
  ) +
  theme(legend.position = "right")
```

## Correlation Matrix

```{r}
#| label: fig-corrplot
#| fig-cap: "Pairwise correlations between the outcome variable and key predictors. Circle size and colour intensity reflect the strength of the correlation."

cor_vars <- country_clean %>%
  select(
    `Full vax %`    = fully_vax_pct,
    `Log GDP`       = log_gdp,
    `HDI`           = hdi,
    `Median age`    = median_age,
    `Aged 65+`      = aged_65_older,
    `Stringency`    = stringency_mean
  ) %>%
  drop_na()

cor_matrix <- cor(cor_vars, use = "complete.obs")

corrplot(cor_matrix,
         method      = "circle",
         type        = "upper",
         tl.col      = "black",
         tl.cex      = 0.85,
         cl.cex      = 0.75,
         col         = colorRampPalette(c("#D55E00", "white", "#009E73"))(200),
         addCoef.col = "black",
         number.cex  = 0.75)
```

# Descriptive Statistics

```{r}
#| label: tbl-descriptive
#| tbl-cap: "Descriptive statistics for key variables by continent."

desc_table <- country_clean %>%
  group_by(Continent = continent) %>%
  summarise(
    N                 = n(),
    `Mean full vax (%)` = round(mean(fully_vax_pct, na.rm = TRUE), 1),
    `Median full vax (%)` = round(median(fully_vax_pct, na.rm = TRUE), 1),
    `SD full vax`     = round(sd(fully_vax_pct, na.rm = TRUE), 1),
    `Mean GDP (USD)`  = round(mean(gdp_per_capita, na.rm = TRUE), 0),
    `Mean HDI`        = round(mean(hdi, na.rm = TRUE), 3),
    `Mean median age` = round(mean(median_age, na.rm = TRUE), 1),
    `Mean stringency` = round(mean(stringency_mean, na.rm = TRUE), 1),
    .groups = "drop"
  )

desc_table %>%
  kable(booktabs = TRUE, align = "lrrrrrrr") %>%
  kable_styling(full_width = TRUE, font_size = 10)
```

# Statistical Modelling

Three OLS regression models are estimated with increasing complexity. Model 1
uses GDP per capita alone as a baseline. Model 2 includes all key predictors.
Model 3 adds continent fixed effects to account for unobserved regional factors.
Heteroscedasticity-consistent (HC3) robust standard errors are reported
throughout.

## Model Estimation

```{r}
#| label: regression-models

m1 <- lm(fully_vax_pct ~ log_gdp,
          data = country_clean)

m2 <- lm(fully_vax_pct ~ log_gdp + hdi + median_age +
            aged_65_older + stringency_mean,
          data = country_clean)

m3 <- lm(fully_vax_pct ~ log_gdp + hdi + median_age +
            aged_65_older + stringency_mean + continent,
          data = country_clean)
```

## Model Summaries

```{r}
#| label: tbl-model1
#| tbl-cap: "Model 1: GDP per capita only."

tidy(m1, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

```{r}
#| label: tbl-model2
#| tbl-cap: "Model 2: Full predictor set, no fixed effects."

tidy(m2, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

```{r}
#| label: tbl-model3
#| tbl-cap: "Model 3: Full predictor set with continent fixed effects."

tidy(m3, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

## Model Fit Comparison

```{r}
#| label: tbl-model-fit
#| tbl-cap: "R-squared and adjusted R-squared for all three models."

tibble(
  Model          = c("M1: GDP only", "M2: Full", "M3: Full + Continent FE"),
  R2             = c(summary(m1)$r.squared,
                     summary(m2)$r.squared,
                     summary(m3)$r.squared),
  `Adj. R2`      = c(summary(m1)$adj.r.squared,
                     summary(m2)$adj.r.squared,
                     summary(m3)$adj.r.squared),
  AIC            = c(AIC(m1), AIC(m2), AIC(m3))
) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

## Multicollinearity Check

```{r}
#| label: tbl-vif
#| tbl-cap: "Variance inflation factors (VIF) for Model 2. Values above 5 indicate problematic multicollinearity."

vif(m2) %>%
  enframe(name = "Variable", value = "VIF") %>%
  mutate(VIF = round(VIF, 3)) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

## Robust Standard Errors (Model 3)

```{r}
#| label: tbl-robust-se
#| tbl-cap: "Model 3 coefficients with HC3 heteroscedasticity-consistent standard errors."

coeftest(m3, vcov = vcovHC(m3, type = "HC3")) %>%
  tidy() %>%
  filter(!str_starts(term, "continent")) %>%
  mutate(
    term = recode(term,
      "(Intercept)"    = "Intercept",
      "log_gdp"        = "Log GDP per capita",
      "hdi"            = "Human Development Index",
      "median_age"     = "Median age",
      "aged_65_older"  = "Population aged 65+",
      "stringency_mean"= "Mean stringency index"
    ),
    across(where(is.numeric), ~ round(., 4))
  ) %>%
  kable(booktabs = TRUE,
        col.names = c("Variable", "Estimate", "Std. Error", "t", "p-value")) %>%
  kable_styling(full_width = FALSE)
```

## Coefficient Plot

```{r}
#| label: fig-coefplot
#| fig-cap: "Regression coefficients with 95% confidence intervals from Model 3 (continent fixed effects not shown). Points to the right of the dashed line indicate a positive association with vaccination rate."

coef_data <- tidy(m3, conf.int = TRUE) %>%
  filter(
    term != "(Intercept)",
    !str_starts(term, "continent")
  ) %>%
  mutate(term = recode(term,
    "log_gdp"         = "Log GDP per capita",
    "hdi"             = "Human Development Index",
    "median_age"      = "Median age",
    "aged_65_older"   = "Population aged 65+",
    "stringency_mean" = "Mean stringency index"
  ))

ggplot(coef_data,
       aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_segment(aes(x = conf.low, xend = conf.high,
                   yend = reorder(term, estimate)),
               colour = "grey40", linewidth = 0.9) +
  geom_point(size = 3.5, shape = 16, colour = "grey20") +
  labs(
    x = "Regression coefficient (95% CI)",
    y = NULL
  )
```

## Regression Diagnostics

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic plots for Model 3."
#| fig-width: 8
#| fig-height: 6

par(mfrow = c(2, 2), family = "sans", cex = 0.8)
plot(m3)
```

# Clustering Analysis

Countries are grouped using k-means clustering on six standardised variables:
full vaccination rate, log GDP, HDI, median age, proportion aged 65+, and mean
stringency. The optimal number of clusters is selected using the elbow method.

```{r}
#| label: clustering-setup

cluster_countries <- country_clean %>%
  filter(!is.na(fully_vax_pct), !is.na(log_gdp), !is.na(hdi),
         !is.na(median_age), !is.na(aged_65_older), !is.na(stringency_mean))

cluster_vars <- cluster_countries %>%
  select(fully_vax_pct, log_gdp, hdi, median_age,
         aged_65_older, stringency_mean)

cluster_scaled <- scale(cluster_vars)
```

## Elbow Plot

```{r}
#| label: fig-elbow
#| fig-cap: "Elbow plot for k-means clustering. The optimal k is identified at the point of diminishing returns in within-cluster sum of squares."

set.seed(42)
wss <- sapply(1:10, function(k) {
  kmeans(cluster_scaled, centers = k, nstart = 25)$tot.withinss
})

ggplot(data.frame(k = 1:10, wss = wss),
       aes(x = k, y = wss)) +
  geom_line(linewidth = 0.8, colour = "grey30") +
  geom_point(size = 3, shape = 16, colour = "grey20") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Number of clusters (k)",
    y = "Total within-cluster sum of squares"
  )
```

## K-Means Solution

```{r}
#| label: kmeans-run

set.seed(42)
km4 <- kmeans(cluster_scaled, centers = 4, nstart = 50)

cluster_countries <- cluster_countries %>%
  mutate(cluster = factor(km4$cluster))
```

## Cluster Profiles

```{r}
#| label: tbl-cluster-profiles
#| tbl-cap: "Profile of each cluster: mean values of key variables."

cluster_profile <- cluster_countries %>%
  group_by(Cluster = cluster) %>%
  summarise(
    N                = n(),
    `Mean full vax (%)`   = round(mean(fully_vax_pct), 1),
    `Mean GDP (USD)` = round(mean(gdp_per_capita, na.rm = TRUE), 0),
    `Mean HDI`       = round(mean(hdi, na.rm = TRUE), 3),
    `Mean age`       = round(mean(median_age, na.rm = TRUE), 1),
    `Mean stringency`= round(mean(stringency_mean, na.rm = TRUE), 1),
    .groups = "drop"
  )

cluster_profile %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```

## Cluster Visualisation (PCA)

```{r}
#| label: fig-pca
#| fig-cap: "PCA biplot of country clusters. Each point represents one country; colour indicates cluster membership and shape indicates continent."

pca_res <- prcomp(cluster_scaled, scale. = FALSE)
pca_df  <- as.data.frame(pca_res$x[, 1:2]) %>%
  mutate(
    cluster   = cluster_countries$cluster,
    continent = cluster_countries$continent,
    country   = cluster_countries$location
  )

var_exp <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

ggplot(pca_df, aes(x = PC1, y = PC2,
                   colour = cluster, shape = continent)) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_colour_manual(
    values = c("1" = "#E69F00", "2" = "#56B4E9",
               "3" = "#009E73", "4" = "#D55E00"),
    name = "Cluster"
  ) +
  scale_shape_manual(values = continent_shapes, name = "Continent") +
  labs(
    x = paste0("PC1 (", var_exp[1], "% variance)"),
    y = paste0("PC2 (", var_exp[2], "% variance)")
  ) +
  theme(legend.position = "right")
```

## Continental Composition of Clusters

```{r}
#| label: fig-cluster-composition
#| fig-cap: "Continental composition of each cluster shown as a stacked horizontal bar chart."

cluster_continent <- cluster_countries %>%
  count(cluster, continent) %>%
  group_by(cluster) %>%
  mutate(pct = n / sum(n) * 100)

ggplot(cluster_continent,
       aes(x = pct, y = cluster, fill = continent)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = continent_colours, name = "Continent") +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  labs(
    x = "Proportion of countries in cluster (%)",
    y = "Cluster"
  ) +
  theme(legend.position = "right")
```

# Comparative Analysis

## Top and Bottom Performers

```{r}
#| label: fig-top-bottom
#| fig-cap: "The 20 highest and 20 lowest performing countries by full vaccination rate. Circles denote top performers; triangles denote bottom performers."
#| fig-height: 9

top20    <- country_clean %>%
  arrange(desc(fully_vax_pct)) %>%
  slice_head(n = 20) %>%
  mutate(group = "Top 20")

bottom20 <- country_clean %>%
  arrange(fully_vax_pct) %>%
  slice_head(n = 20) %>%
  mutate(group = "Bottom 20")

top_bottom <- bind_rows(top20, bottom20)

ggplot(top_bottom,
       aes(x = fully_vax_pct,
           y = reorder(location, fully_vax_pct),
           shape = group, colour = group)) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Top 20"    = "#009E73",
                                 "Bottom 20" = "#D55E00"),
                      name = NULL) +
  scale_shape_manual(values  = c("Top 20"    = 16,
                                 "Bottom 20" = 17),
                     name = NULL) +
  scale_x_continuous(labels = label_percent(scale = 1),
                     limits = c(0, 100)) +
  labs(
    x = "Population fully vaccinated (%)",
    y = NULL
  ) +
  theme(
    legend.position = "top",
    axis.text.y     = element_text(size = 7)
  )
```

## Summary: Top vs Bottom 20

```{r}
#| label: tbl-top-bottom
#| tbl-cap: "Mean characteristics of the top 20 and bottom 20 countries by vaccination rate."

top_bottom %>%
  group_by(Group = group) %>%
  summarise(
    `Mean full vax (%)` = round(mean(fully_vax_pct), 1),
    `Mean GDP (USD)`    = round(mean(gdp_per_capita, na.rm = TRUE), 0),
    `Mean HDI`          = round(mean(hdi, na.rm = TRUE), 3),
    `Mean median age`   = round(mean(median_age, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  kable(booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```


```{r}
#| label: session-info
#| code-fold: false

sessionInfo()
```
